Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-18 writing/paper-rbm/letters/cover_letter.Rmd
ready for submission

---
author: 
 - Andee Kaplan
 - Postdoctoral Associate
 - Department of Statistical Sciences 
 - Duke University
date: "`r format(Sys.time(), '%B %d, %Y')`"
opening: Dear Drs. Liu and Zhao,
closing: Best regards,
address: 
 - Regina Liu and Hongyu Zhao
 - Co-Editors-Elect
 - "*Journal of the American Statistical Association - Theory & Methods*"
return-address:
 - Andee Kaplan
 - Department of Statistical Sciences
 - Duke University
 - P.O. Box 90251
 - Durham, NC 27708-0251
output:
  pdf_document:
    template: template-letter.tex
geometry: margin=1in
---

I am writing to submit an article entitled "Properties and Bayesian fitting of restricted Boltzmann machines" for consideration for publication in the Journal of the American Statistical Association - Theory & Methods.  In this submission, we discuss the relationship between parameter specification and unattractive model properties---degeneracy, instability and uninterpretability---in restricted Boltzmann machines and describe related difficulties that can arise in likelihood-based and Bayes fitting of such models.

This paper references results from another manuscript, an invited revision of which is presently under review by Technometrics. We have provided a blinded copy of this second manuscript as a supplement for referees.

Thank you for your consideration.

>>>>

---
author: 
 - Andee Kaplan
 - Postdoctoral Associate
 - Department of Statistical Sciences 
 - Duke University
date: "`r format(Sys.time(), '%B %d, %Y')`"
opening: Dear Dr. Clarke,
closing: Best regards,
address: 
 - Dr. Bertrand Clarke
 - Editor
 - "Statistical Analysis and Data Mining"
return-address:
 - Andee Kaplan
 - Department of Statistical Sciences
 - Duke University
 - P.O. Box 90251
 - Durham, NC 27708-0251
output:
  pdf_document:
    template: template-letter.tex
geometry: margin=1in
---

My name is Andee Kaplan, and I am a postdoctoral fellow at Duke University in the Department of Statistical Science. I am writing on behalf of my co-authors, Dr. Daniel Nordman and Dr. Stephen Vardeman, to submit our article entitled "Properties and Bayesian fitting of restricted Boltzmann machines" for consideration for publication in Statistical Analysis and Data Mining: the ASA Data Mining Journal.

In this submission, we discuss the relationship between parameter specification and unattractive model properties---degeneracy, instability and uninterpretability---in restricted Boltzmann machines and describe related difficulties that can arise in likelihood-based and Bayes fitting of such models.  Our findings are that these models can experience certain types of model instability problems over the parameter space, which relates to how useful the model may be for representing realistic data-generation. In addition to the concepts and theories of instability, we also present empirical evidence that if parameter values stray too far from the origin of the parameter space, then these model problems can arise. We also investigate a principled Bayes fitting method with the aim to a) potentially quantify the uncertainty in estimation and b) potentially avoid problem areas of the parameter space by constraining parameters appropriately. Our methods show promise, but are numerically intractable in large data cases. In our estimation, it is unclear that current practice with these models is addressing our presented concerns.

In light of this, we do not expect this message to be well-received by proponents of the current methodology, and respectfully request that the review of the paper be carried out with this in mind and involve reviewers that have no personal stake in the existing practice.

This paper references results from another manuscript, a paper which is presently under review by Information and Inference: A Journal of the IMA. We have provided a copy of this second manuscript as a supplement for referees.

Thank you for your time and consideration, and do not hesitate to contact me if you have any questions regarding our paper.

### 2018-09-18 writing/paper-rbm/paper.tex
ready for submission

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}

@@

\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}

@@

\begin{figure}[ht]
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{../resources/images/tikz/toyhull_top.tikz}

    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{120}
    \input{../resources/images/tikz/toyhull_top.tikz}

@@

{\centering \includegraphics{paper_files/figure-latex/degen-toy-1}

@@

{\centering \includegraphics{paper_files/figure-latex/instab-1}

@@

{\centering \includegraphics{paper_files/figure-latex/uninterp-1}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/degen-plots-1.pdf}
\caption{\label{fig:degen-plots}Results from the numerical experiment,
here looking at the fraction of models that were near-degenerate for
each combination of magnitudes of the points \(\boldsymbol \theta\) and
model size, where
\(\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})\)
is split into main and interaction parameters. Black lines show the
contour levels for fraction of near-degeneracy, while the thick black
line shows the level where the fraction of near-degenerate models is
.05.}
\end{figure}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/instab-plots-1.pdf}
\caption{\label{fig:instab-plots}The sample mean value of
\(\text{LREP}(\boldsymbol \theta)/\nv\) at each grid point for each
combination of magnitudes of the points of \(\boldsymbol \theta\) and
model size. As the magnitude of \(\boldsymbol \theta\) grows, so does
the value of this metric, indicating typical instability in the model.}
\end{figure}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/uninterp-plots-1.pdf}
\caption{\label{fig:uninterp-plots}The sample mean of the maximum
component of the absolute difference between the model expectation
vector, E\(\left[\boldsymbol X | \boldsymbol \theta\right]\), and the
expectation vector given independence,
E\(\left[\boldsymbol X | \boldsymbol \theta^* \right ]\). Larger
magnitudes of \(\boldsymbol \theta\) correspond to larger differences,
thus indicating reduced interpretability.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/split-plots-dist-1.pdf}
\caption{\label{fig:split-plots-dist}The fraction of near-degenerate
models for each magnitude of \(\boldsymbol \theta\). For each number
\(\nv\) of visibles in the model, the fraction near-degenerate moves
away from zero at larger values of \(||\boldsymbol \theta||\) as the
number \(\nh\) of hidden variables increases and the slope becomes
steeper as \(\nh\) increases as well.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/acf-1.pdf}
\caption{\label{fig:acf}The autocorrelation functions (ACF) for the
posterior probabilityies of all \(2^4 = 16\) possible outcomes for the
vector of four visibles assessed at multiple lags for each method with
BwTNLV in black and BwTNML in red. As expected, ACF corresponding to the
method that marginalizes out the hidden variables from the likelihood
decreases to zero at a much faster rate, indicating better mixing for
the chain.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/fitting-plot-1.pdf}
\caption{\label{fig:fitting-plot}Posterior probabilities of \(16 = 2^4\)
possible realizations of \(4\) visibles using two of the three Bayesian
fitting techniques, BwTPLV and BwTNML. The black vertical lines show the
true probabilities of each vector of visibles based on the parameters
used to generate the training data while the red vertical lines show the
empirical distribution. BwTNML produces the best fit for the data,
however is also the most computationally intensive and least feasible
with a real dataset.}
\end{figure}

>>>>

\begin{figure}
\includegraphics[width=1\linewidth]{sample-models-1} \includegraphics[width=1\linewidth]{sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}

@@

\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}

@@

\begin{figure}[ht]
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{toyhull_top.tikz}

    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{120}
    \input{toyhull_top.tikz}

@@

{\centering \includegraphics{degen-toy-1}

@@

{\centering \includegraphics{instab-1}

@@

{\centering \includegraphics{uninterp-1}

@@

\begin{figure}
\centering
\includegraphics{degen-plots-1.eps}
\caption{\label{fig:degen-plots}Results from the numerical experiment,
here looking at the fraction of models that were near-degenerate for
each combination of magnitudes of the points \(\boldsymbol \theta\) and
model size, where
\(\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})\)
is split into main and interaction parameters. Black lines show the
contour levels for fraction of near-degeneracy, while the thick black
line shows the level where the fraction of near-degenerate models is
.05.}
\end{figure}

\begin{figure}
\centering
\includegraphics{instab-plots-1.eps}
\caption{\label{fig:instab-plots}The sample mean value of
\(\text{LREP}(\boldsymbol \theta)/\nv\) at each grid point for each
combination of magnitudes of the points of \(\boldsymbol \theta\) and
model size. As the magnitude of \(\boldsymbol \theta\) grows, so does
the value of this metric, indicating typical instability in the model.}
\end{figure}

\begin{figure}
\centering
\includegraphics{uninterp-plots-1.eps}
\caption{\label{fig:uninterp-plots}The sample mean of the maximum
component of the absolute difference between the model expectation
vector, E\(\left[\boldsymbol X | \boldsymbol \theta\right]\), and the
expectation vector given independence,
E\(\left[\boldsymbol X | \boldsymbol \theta^* \right ]\). Larger
magnitudes of \(\boldsymbol \theta\) correspond to larger differences,
thus indicating reduced interpretability.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{split-plots-dist-1.eps}
\caption{\label{fig:split-plots-dist}The fraction of near-degenerate
models for each magnitude of \(\boldsymbol \theta\). For each number
\(\nv\) of visibles in the model, the fraction near-degenerate moves
away from zero at larger values of \(||\boldsymbol \theta||\) as the
number \(\nh\) of hidden variables increases and the slope becomes
steeper as \(\nh\) increases as well.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{acf-1.eps}
\caption{\label{fig:acf}The autocorrelation functions (ACF) for the
posterior probabilityies of all \(2^4 = 16\) possible outcomes for the
vector of four visibles assessed at multiple lags for each method with
BwTNLV in black and BwTNML in red. As expected, ACF corresponding to the
method that marginalizes out the hidden variables from the likelihood
decreases to zero at a much faster rate, indicating better mixing for
the chain.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{fitting-plot-1.eps}
\caption{\label{fig:fitting-plot}Posterior probabilities of \(16 = 2^4\)
possible realizations of \(4\) visibles using two of the three Bayesian
fitting techniques, BwTPLV and BwTNML. The black vertical lines show the
true probabilities of each vector of visibles based on the parameters
used to generate the training data while the red vertical lines show the
empirical distribution. BwTNML produces the best fit for the data,
however is also the most computationally intensive and least feasible
with a real dataset.}
\end{figure}
